---
title: "Student Parental Education Analysis"
author: "Team 24"
date: "2024-12-03"
output: html_document
---

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(car)


data<-read.csv("StudentsPerformance.csv")
head(data)
```

```{r}
data <- data %>%
  mutate(parental.level.of.education = case_when(
    parental.level.of.education == "associate's degree" ~ "Assoc. Deg.",
    parental.level.of.education == "bachelor's degree" ~ "Bach. Deg.",
    parental.level.of.education == "master's degree" ~ "Mast. Deg.",
    parental.level.of.education == "some college" ~ "Some College",
    parental.level.of.education == "high school" ~ "High School",
    parental.level.of.education == "some high school" ~ "Some HS",
    TRUE ~ parental.level.of.education # Leave unchanged if no match
  ))


# Reshape data to a long format for separate histograms
data_long_categorical <- data %>%
  select(parental.level.of.education) %>%
  pivot_longer(cols = everything(), names_to = "category", values_to = "value")

# Create a combined histogram figure with facets
ggplot(data_long_categorical, aes(x = value)) +
  geom_bar(fill = "#4B4B4B") +  # Unified color for all histograms
  facet_wrap(~ category, scales = "free_x", ncol = 3) +  # Layout with 3 columns to center bottom two
  labs(
    title = "Distribution of Education Levels",
    x = "Category Value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "none"
  )
```


non education
```{r}
# Select non-parental education factors
non_parental_factors <- data %>%
  select(gender, race.ethnicity, lunch, test.preparation.course)

# Reshape data to long format for plotting
non_parental_long <- non_parental_factors %>%
  pivot_longer(cols = everything(), names_to = "Category", values_to = "Value")

# Create a bar chart for distribution of non-parental factors
ggplot(non_parental_long, aes(x = Value)) +
  geom_bar(fill = "#4B4B4B") +
  facet_wrap(~ Category, scales = "free", ncol = 2) +
  labs(
    title = "Distribution of Non-Parental Education Factors",
    x = "Categories",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
# Calculate the average score for each student
data <- data %>%
  mutate(average.score = rowMeans(select(., math.score, reading.score, writing.score)))

# Create a boxplot for average scores
ggplot(data, aes(x = parental.level.of.education, y = average.score)) +
  geom_boxplot(fill = "#4B4B4B", color = "black", alpha = 0.7) +
  labs(
    title = "Average Scores by Parental Education",
    x = "Parental Education Level",
    y = "Average Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Check for homogenity of variances before continuing with anova
```{r}
levene_average<-leveneTest(average.score ~ parental.level.of.education, data = data)
levene_average
```

We can continue with anova

Anova Average
```{r}
anova_average <- aov(average.score ~ parental.level.of.education, data = data)
summary(anova_average)
```

box plots for all scores 

```{r}
scores <- c("math.score", "reading.score", "writing.score")
titles <- c("Math Scores by Parental Education", "Reading Scores by Parental Education", "Writing Scores by Parental Education")
# Loop through each score and display the boxplot
for (i in seq_along(scores)) {
  p <- ggplot(data, aes(x = parental.level.of.education, y = .data[[scores[i]]])) +
    geom_boxplot(fill = "#4B4B4B", color = "black", alpha = 0.7) +
    labs(
      title = titles[i],
      x = "Parental Education Level",
      y = "Score"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 16),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  print(p) # Correctly print the plot within the loop
}
```
levene for math scores

```{r}
levene_math<-leveneTest(math.score ~ parental.level.of.education, data = data)
levene_math
```
Anova for math and education level

```{r}
anova_math <- aov(math.score ~ parental.level.of.education, data = data)
summary(anova_math)
```

check for homogeneity of variance across groups. Levene test
```{r}
levene_read<-leveneTest(reading.score ~ parental.level.of.education, data = data)
levene_read
```

Anova for Reading and education level

```{r}
anova_reading <- aov(reading.score ~ parental.level.of.education, data = data)
summary(anova_reading)
```

Check for homogenity of variances before continuing with anova
```{r}
levene_writing<-leveneTest(writing.score ~ parental.level.of.education, data = data)
levene_writing
```

Anova for Writing and education level

```{r}
anova_write <- aov(writing.score ~ parental.level.of.education, data = data)
summary(anova_write)
```


GLM Univariate

```{r}
glm_average <- glm(
  average.score ~ parental.level.of.education,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(glm_average)
```


GLM Math
```{r}
data$parental.level.of.education <- factor(
  data$parental.level.of.education,
  levels = c("High School", "Some HS", "Some College", "Assoc. Deg.", "Bach. Deg.", "Mast. Deg.") # Customize the order
)

glm_math <- glm(
  math.score ~ parental.level.of.education,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(glm_math)
```



GLM Reading
```{r}
glm_read <- glm(
  reading.score ~ parental.level.of.education,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(glm_read)
```


GLM Writing
```{r}
glm_write <- glm(
  writing.score ~ parental.level.of.education,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(glm_write)
```


Bivariate Average results
```{r}
bi_glm_average_gender <- glm(
  average.score ~ parental.level.of.education + gender,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(bi_glm_average_gender)
```


```{r}
bi_glm_average_race <- glm(
  average.score ~ parental.level.of.education + race.ethnicity,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(bi_glm_average_race)
```

```{r}
bi_glm_average_lunch <- glm(
  average.score ~ parental.level.of.education + lunch,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(bi_glm_average_lunch)
```



```{r}
bi_glm_average_test <- glm(
  average.score ~ parental.level.of.education + test.preparation.course,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(bi_glm_average_test)
```


Full Model

```{r}
full_glm_average <- glm(
  average.score ~ parental.level.of.education + test.preparation.course+lunch + race.ethnicity+gender,
  data = data,
  family = gaussian() # Assumes a normal distribution for the dependent variable
)
summary(full_glm_average)
```

vif
```{r}
vif_full<-vif(full_glm_average)
vif_full
```
No evidence of multicollinearity


Now lets use stepwise model selection to find the best fit model

```{r}
library(MASS)

best_model <- stepAIC(full_glm_average, direction = "both", trace = TRUE)
summary(best_model)
```
Looks like parental level of education remains a significant predictor alongside other predictors of test score.
